---
title: "CUSTOMER BRAND PREFERNCE"
author: "Victor Sizah"
goal: "PREDICTING CUSTOMER BRAND PREFERENCE"
date: "2024-03-18"
output: html_document
---

```{r setup, include=FALSE}

```

```{r warning=FALSE, message=FALSE, include=FALSE}
library(knitr)
library(lattice)
library(ggplot2)
library(caret)
library(pls)
library(dplyr)
library(tidyr)
library(kernlab)
library(plotly)
library(stats)
library(formattable)
```

# INTRODUCTION

**GOALS OF THE PROJECT**
  The goal of this project is to predict the most preferred brand
  
**THE VARIOUS HYPOTHESIS FORMULATED INCLUDE:**
  1.Individuals with higher levels of education like Master's, Doctoral, or Professional     Degree, might have higher salaries compared to those with lower levels of education.
  2.Individuals with higher incomes might be more likely to own luxury car brands like        Mercedes Benz or Cadillac.
  3.Older individuals might have higher credit scores due to longer credit histories and        more stable financial situations.
  4.Older individuals might have higher salaries due to more experience and higher             positions in their careers.

# CHAPTER 2

  **DATA DESCRIPTION**
```{r}
My_Data <- read.csv("CompleteResponses.csv ")
```

```{r}
summary(My_Data)
```
```{r}
new_My_Data <- My_Data
```

```{r}
str(new_My_Data)
```
**Checking datatypes**
```{r}
elevel <- new_My_Data$elevel

if (is.numeric(elevel)) {
  print("The column is numeric.")
} else if (is.factor(elevel)) {
  print("The column is a factor.")
} else {
  print("The column is neither numeric nor a factor.")
}
```

**CONVERTING NUMERIC DATA COLUMN INTO CATEGORICAL DATA**
```{r}
new_My_Data$elevel <- as.factor(new_My_Data$elevel)
new_My_Data$car <- as.factor(new_My_Data$car)
new_My_Data$zipcode <- as.factor(new_My_Data$zipcode)
new_My_Data$brand <- as.factor(new_My_Data$brand)

# Checking the converted column
str(new_My_Data$elevel)  
str(new_My_Data$car) 
str(new_My_Data$zipcode) 
str(new_My_Data$brand) 

```

**MAPPING ALL THE CATEGORICAL VARIABLES**
```{r}
# Define a named vector for mapping
factor_to_elevel <- c("0" = "Less than High School Degree", "1" = "High School Degree", "2" = "Some College", "3" = "4-year College degree", "4" = "Master's, Doctoral or Professional Degree")

# Replace numeric values with corresponding words
new_My_Data$elevel <- factor_to_elevel[as.character(new_My_Data$elevel)]
```

```{r}
factor_to_car <- c("1" = "BMW", "2" =	"Buick", "3" =	"Cadillac", "4" =	"Chevrolet", "5"	= "Chrysler", "6"	= "Dodge", "7" =	"Ford", "8"	= "Honda", "9" =	"Hyundai", "10" = "Jeep", "11" =	"Kia", "12" =	"Lincoln", "13"	= "Mazda", "14"	= "Mercedes Benz", "15"	= "Mitsubishi", "16" =	"Nissan", "17"	= "Ram", "18" =	"Subaru", "19" =	"Toyota", "20" = "None of the above")
new_My_Data$car <- factor_to_car[as.character(new_My_Data$car)]
```

```{r}
factor_to_zipcode <- c("0" = "New England", "1" =	"Mid-Atlantic", "2" =	"East North Central", "3" =	"West North Central", "4" =	"South Atlantic", "5" =	"East South Central", "6" =	"West South Central", "7" =	"Mountain", "8" =	"Pacific")
new_My_Data$zipcode <- factor_to_zipcode[as.character(new_My_Data$zipcode)]
```

```{r}
factor_to_brand <- c("0" = "Acer", "1"	= "Sony")
new_My_Data$brand <- factor_to_brand[as.character(new_My_Data$brand)]
```
```{r}
new_My_Data$brand <- as.factor(new_My_Data$brand)
```



```{r}
summary(new_My_Data)
```
# Data Preparation

**check missing values in all attributes**

```{r}
missing_values <- colSums(is.na(new_My_Data))
print(missing_values)
```
There are no MISSING VALUES in our data set


**Checking for outliers (using a box plot and its only used for numeric datatype)**

```{r}
boxplot(new_My_Data$salary, main = "Boxplot of Salary")
boxplot(new_My_Data$age, main = "Boxplot of Age")
boxplot(new_My_Data$credit, main = "Boxplot of Credit")
```
There are no OUTLIERS in the dataset

**FEATURE ENGINEERING**
creating a new variable age-range using the age attribute(feature engineering)(research on data rangling )

```{r}
new_My_Data$age_range <- cut(new_My_Data$age, breaks = c(0, 18, 30, 50, 100), labels = c("0-18", "19-30", "31-50", "51-100"))
```


**Checking for inconsistences in the data**

```{r}
summary(new_My_Data)
```


**Renaming the column to suit what they represent (elevel= educational_level, credit= amount_credit)**

```{r}
names(new_My_Data)[names(new_My_Data) == "elevel"] <- "educational_level"
names(new_My_Data)[names(new_My_Data) == "credit"] <- "amount_credit"

```


**checking for duplicate**

```{r}
Duplicates <- any(duplicated(new_My_Data))

if (Duplicates) {
  print("The dataframe contains duplicates.")
} else {
  print("The dataframe does not contain duplicates.")
}

```
```{r}
Duplicates <- new_My_Data[duplicated(new_My_Data), ] 
print (Duplicates)
```

I OBSERVED THAT THERE IS NOT DUPLICATE

**checking for zero variance**

```{r}
salary <- new_My_Data$salary

if (var(salary) == 0) {
  print("The column has zero variance (all values are the same).")
} else {
  print("The column does not have zero variance.")
}
```
```{r}
age <- new_My_Data$age

if (var(age) == 0) {
  print("The column has zero variance (all values are the same).")
} else {
  print("The column does not have zero variance.")
}
```

```{r}
amount_credit <- new_My_Data$amount_credit

if (var(amount_credit) == 0) {
  print("The column has zero variance (all values are the same).")
} else {
  print("The column does not have zero variance.")
}
```
```{r}
brand <- My_Data$brand
variance <- apply(My_Data, 2, var)

if (var(brand) == 0) {
  print("The column has zero variance (all values are the same).")
} else {
  print("The column does not have zero variance.")
}
```
**correlation matrix**

```{r}
correlation_matrix <- cor(My_Data)
print(correlation_matrix)
```



# EDA

# UNIVARIANT ANALYSIS


**Histogram distribution of Salary**
```{r}
ggplot(new_My_Data, aes(x=salary)) + geom_histogram(color="darkblue", fill="lightblue", bin=20) +labs(x = "Salary Earned", title = "A Histogram of salary", subtitle = "Showing Salary with 20 Bins",
 caption ="Salary- Count",                                                                                  alt = "salary- Count in frequency")
```


**Histogram distribution of Brand**
```{r}
ggplot(My_Data, aes(x=brand)) + geom_histogram(color="darkblue", fill="lightblue") +labs(x = "Brand", title = "A Histogram of Brand", subtitle = "",
 caption ="brand- Count",                                                                                  alt = "salary- Count in frequency")
```

# BIVARIANT

**Distribution between educational_level and Salary**
```{r}
plot1 <- ggplot(new_My_Data, aes(salary, educational_level))
plot1 + geom_col()
```

**Distribution between Salary and Car**
```{r}
plot1 <- ggplot(new_My_Data, aes(salary, car))
plot1 + geom_col()
```


```{r}
cq <-ggplot(new_My_Data, aes(age))
cq +geom_area(stat= "bin")
```

```{r}
cq +geom_density(kernel= "gaussian")
```

```{r}
cq +geom_freqpoly(kernel= "gaussian")
```

```{r}
cq +geom_histogram(binwidth = 5)
```
```{r}
cr <- ggplot(new_My_Data)
cr +geom_qq(aes(sample = age))
```

**Density plot of salary**
```{r}
c1 <-ggplot(new_My_Data, aes(salary))
c1+geom_density(kernel= "gaussian")
```
**Density plot of Age**
```{r}
c2 <-ggplot(new_My_Data, aes(age))
c2+geom_density(kernel= "gaussian")
```
```{r}
c1 <-ggplot(new_My_Data, aes(amount_credit))
c1+geom_density(kernel= "gaussian")
```

**A linear regression model developed to predict salary based on age, educational level, car ownership, zipcode, credit score, and brand**
```{r}
model <- lm(salary ~ age + elevel + car + zipcode + credit + brand, data = My_Data)
# View summary of the regression model
summary(model)

```
**Bar plot illustrates the distribution of brands in the dataset.**
```{r}
d <- ggplot(new_My_Data, aes(brand))
d + geom_bar()
```

**Bar plot illustrates the distribution of education_level in the dataset.**
```{r}
d1 <- ggplot(new_My_Data, aes(educational_level))
d1 + geom_bar() + coord_flip()
```
**Bar plot illustrates the distribution of car in the dataset.**
```{r}
d2 <- ggplot(new_My_Data, aes(car))
d2 + geom_bar() + coord_flip()
```

**Jitter plot to show the distribution of salary against amount of credit**
```{r}
e <- ggplot(new_My_Data, aes(salary, amount_credit))
e + geom_jitter(height = 2, width = 2) 
```

**Smooth plot to show the distribution of salary against amount of credit**
```{r}
e1 <- ggplot(new_My_Data, aes(salary, amount_credit))
e1 + geom_smooth()
```

**Point plot to show the distribution of salary against age**
```{r}
e2 <- ggplot(new_My_Data, aes(salary, age))
e2 + geom_point()
```
**Column plot to show the distribution of salary against age**
```{r}
f <- ggplot(new_My_Data, aes(educational_level, salary))
f + geom_col() + coord_flip()
```
**Violin plot to show the distribution of salary against age**
```{r}
f1 <- ggplot(new_My_Data, aes( brand, salary))
f1  + geom_violin(scale = "area")
```
**Histogram plot to show the relationship between salary and brand**
```{r}
#HISTOGRAM OF SALARY AND BRAND
ggplot (new_My_Data, aes(x=salary, fill=brand)) +
geom_histogram (color="black", bins=10) +
labs (title="Relationship Between Brand and Salary") + 
scale_x_continuous (breaks=seq(20000,150000, 15000))+
theme (axis.text.x = element_text(angle=60, hjust=1)) +
scale_y_continuous (labels = scales::percent)
```

# MULTIVARIANT ANALYSIS

**Relationship between Age, Salary and Brand**
```{r}
ggplot (new_My_Data,aes (x=age, y=salary, col=brand)) +
geom_point() +
geom_smooth() +
  labs( X = 'Age', Y = "Salary", title= "Multivariate Relationship between Age, salary and Brand")

```

**Box plot between Salary and Brand**
```{r}
ggplot(new_My_Data, aes(x=brand, y=salary, fill=brand)) +
geom_boxplot()+labs(x= "salary", title = "A Boxplot of Salary and Brand",
subtitle = "Showing salary with 20 categories", caption = "Salary to Brand",
alt = "Salary to brand- Count in frequency")+ stat_summary(fun.y = median,
colour="black", geom = "text", vjust= -0.7, aes(label=round(..y.., digits=
1)))
```

**Historgram showing the relation between salary and educational level**
```{r}
ggplot(My_Data, aes(x=salary)) + geom_histogram(color="darkblue",
fill="lightblue", bins=20) + facet_wrap(~elevel, scales = "free_x")
  labs (x="salary", title = "A Histogram of salary and eductaion level",
subtitle = "showing salary with 20 categories", caption = "salary to
eduction level -count", alt = "salary to educational level- count in
frequency")
```

**Splitting the data into a training set and a testing set, with 75% allocated for training, ensuring reproducibility by setting the seed.**
```{r}
set.seed(107)
inTrain <- createDataPartition(y = new_My_Data$brand, p = .75, list = FALSE)
```

**Showing the structured summary of the object 'inTrain'**
```{r}
str(inTrain)
```
```{r}
training <- new_My_Data[inTrain, ]
```

```{r}
testing <- new_My_Data[-inTrain, ]
```

```{r}
nrow(training)
```
```{r}
nrow(testing)
```
**Checking for null values**
```{r}
#is.na(new_My_Data)
if (any(is.na(new_My_Data))) {
  print("There are null values in the dataset.")
} else {
  print("There are no null values in the dataset.")
}
```

# TRAINING MODELS 

**Initializing train control object for cross-validation resampling method**
```{r}
ctrl <- trainControl(method = 'cv')

```


**Training a Partial Least Squares (PLS) regression model to predict the brand variable, optimizing model performance through resampling and tuning**
```{r}
plsFit <- train(
  brand ~ .,
  data = training,
  method = "pls",
  #preproc = c("center", "scale"),
  tuneLength = 15,
  verbose = FALSE,
  ## added:
  trcontrol = ctrl,
  verbosity = 0
)
```
```{r}
plsFit
```
**KNN**
```{r}
knnFit <- train(
  brand ~ .,
  data = training,
  method = "knn"
)
```


```{r}
knnFit
```
**C5.0 MODEL**
```{r}

c5Fit <- train(
  brand ~ .,
  data = training,
  method = "C5.0"
)
```

```{r}
c5Fit
```

**SVMLINEAR MODEL**
```{r}
SvmLinearFit <- train(
  brand ~ .,
  data = training,
  method = "svmLinear"
)
```

```{r}
SvmLinearFit

```

**GBM MODEL**
```{r}
GbmFit <- train(
  brand ~ .,
  data = training,
  method = "gbm"
)
```


**Data validation**

Predictions for the 'brand' variable using the trained C5.0 model on a test dataset.
```{r}
brand_predict <- predict(c5Fit, testing)
```

```{r}
summary(brand_predict)
```

```{r}
summary (testing$brand)
```

**confusion matrix**
```{r}
#creating confusion matrix
c50_confusion_matrix_Brand <- confusionMatrix(data = brand_predict, reference = testing$brand )
c50_confusion_matrix_Brand
```

**Resampling**
```{r}
#Resampling for the best performing model
resamps_Brand <- resamples(list(Knn = knnFit, svmLinear = SvmLinearFit, c5.0 = c5Fit, gbm = GbmFit, pls = plsFit))
summary(resamps_Brand)
```


The best performing model according to accuracy is C50

**Tuning Grid**
```{r warning=FALSE}
#Define the tuning grid
tuneGrid <- expand.grid(.trials = 20, 
                        .model = "tree",
                        .winnow = TRUE)
#Train the model with parameter tuning
c50Fit_Final <- train(
  brand ~ .,
  data = training,
  method = "C5.0",
  tuneGrid = tuneGrid,
  trcontrol = trainControl(method = "cv", number = 5)
)
```

**predicting brand variable using a trained C5.0 model on the testing dataset.**
```{r}
brand_predict <- predict(c50Fit_Final, testing)
```


**Showing the distribution and characteristics of the predicted brand values.**
```{r}
summary(brand_predict)
```

**The confusion matrix summarizes the performance of the C5.0 model by comparing predicted and actual brand values on the test dataset.**
```{r}
c50_confusion_matrix_Brand <- confusionMatrix(data = brand_predict, reference = testing$brand )
c50_confusion_matrix_Brand
```

**Loading new_Survey_Data**
```{r}
load(file= "C:/Users/sizah/Downloads/DA/RProject/new_Survey_Data.rdata")
```


```{r}
test_predict <- predict(c50Fit_Final, new_Survey_Data)
```

```{r}
summary(test_predict)
```

```{r}
new_Survey_Data <- cbind(new_Survey_Data, test_predict)
```

**LOADING DATA FILE INTO ENVIRONMENT**
```{r}
load(file= "C:/Users/sizah/Downloads/DA/RProject/new_Survey_Data_Brand.rdata")
```

**Adding a new column containing brand data to the new_survey_Data dataset.**
```{r}
new_Survey_Data <- cbind(new_Survey_Data, new_Survey_Data_Brand$brand)
```

**The bar plot illustrates the distribution of 'test_predict' levels in the dataset.**
```{r}
d <- ggplot(new_Survey_Data_Brand, aes(test_predict))
d + geom_bar()
```

