---
title: "PROJECT"
author: "SizahVick"
date: "2024-05-10"
output:
  html_document:
    highlight: tango
    theme: cosmo
    toc: yes
    toc_float: TRUE
    toc_depth: 4
    code_folding: hide
  slidy_presentation:
    highlight: tango
goal: 'CUSTOMER SPENDING BEHAVIOR ANALYSIS '
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**LOADING LIBRARIES**
```{r warning=FALSE, message=FALSE, include=FALSE}
library(dplyr)
library(ggplot2)
library(caret)
library(randomForest)

```

# Introduction

**This report presents an analysis of the Demographic Data, focusing on various aspects such as data preparation, data visualization, and model building.**

**The primary goal is to understand customer spending behavior, detect any potential issues like missing values and duplicates, and build predictive models to derive insights.**

# DATA Description
 **LOADING MY DATA Demographic_Data.csv, I OBSERVED THAT THE FILE CONSIST OF 80000 OBJECTS AND 5 VARIABLES IN WHICH THREE (3) OF THE VARIABLES ARE NUMERICAL (AGE, ITEMS AND AMOUNT) AND THE REMAINING TWO (2) VARIABLES ARE CATEGORICAL (IN.STORE AND REGION). **

## Loading the Data
```{r}
Demographic_Data<- read.csv("Demographic_Data.csv")
```

**Summary Statistics**
```{r}
summary(Demographic_Data)
```

**LOADING THE FUNCTION STR IT DISPLAYS THE STRUCTURE OF MY DATA SET**
```{r}
str(Demographic_Data)
```


```{r}
Demographic_Data_A <- Demographic_Data
Demographic_Data_B <- Demographic_Data
new_Demographic_Data <- Demographic_Data
```

```{r}
str(new_Demographic_Data)
```


# DATA PREPARATION

## Checking missing values in all attributes
```{r}
missing_values <- colSums(is.na(new_Demographic_Data))
print(missing_values)
```
**There are no MISSING VALUES in the data set**

## Checking for outliers
```{r}
boxplot(new_Demographic_Data$age, main = "Boxplot of Age")
boxplot(new_Demographic_Data$items, main = "Boxplot of Items")
boxplot(new_Demographic_Data$amount, main = "Boxplot of Amount")
```
**I OBSERVED THAT THERE ARE OUTLINERS IN MY DATA SET**

```{r}
summary(new_Demographic_Data)
```


## Checking for duplicates
```{r}
Duplicates <- any(duplicated(new_Demographic_Data))

if (Duplicates) {
  print("The dataframe contains duplicates.")
} else {
  print("The dataframe does not contain duplicates.")
}
```


```{r}
Duplicates <- new_Demographic_Data[duplicated(new_Demographic_Data), ] 
print (Duplicates)
```
**I OBSERVED THAT THERE ARE DUPLICATES IN MY DATA SET**

```{r}
#Removing duplicates
new_Demographic_Data <- new_Demographic_Data %>%
  distinct()
Demographic_Data_A <- new_Demographic_Data %>%
  distinct()
Demographic_Data_B <- new_Demographic_Data %>%
  distinct()
```


## DATA TRANSFORMATION

**CONVERTING NUMERIC DATA COLUMN INTO CATEGORICAL DATA (The in.store and region columns are converted to factors.)**
```{r}
new_Demographic_Data$in.store <- as.factor(new_Demographic_Data$in.store)
new_Demographic_Data$region <- as.factor(new_Demographic_Data$region)

```

```{r}
# Checking the converted column
str(new_Demographic_Data$in.store)  
str(new_Demographic_Data$region) 
```

**MAPPING ALL THE CATEGORICAL VARIABLES**
```{r}
# Define a named vector for mapping
factor_to_instore <- c("0" = "Online purchases", "1" = "In-store purchases")
# Replace numeric values with corresponding words
new_Demographic_Data$in.store <- factor_to_instore[as.character(new_Demographic_Data$in.store)]
```

```{r}
factor_to_region <- c("1" = "North", "2" =	"South", "3" =	"East", "4" =	"West")
new_Demographic_Data$region <- factor_to_region[as.character(new_Demographic_Data$region)]
```

#DATA ANALYSIS

# 1
**Customers' Spending by Region**
**Customers in different regions that spend more per transaction**
```{r}
# Calculating Average Transaction Amount by Region
avg_transaction_by_region <- aggregate(amount ~ region, data = new_Demographic_Data, FUN = mean)
print(avg_transaction_by_region)
```

**Regions that spend the most and least**
```{r}
# Identifying Regions with Highest and Lowest Spending
region_with_highest_spending <- avg_transaction_by_region[which.max(avg_transaction_by_region$amount), ]
region_with_lowest_spending <- avg_transaction_by_region[which.min(avg_transaction_by_region$amount), ]

print(paste("Region with highest spending:", region_with_highest_spending$region, "-> ($", round(region_with_highest_spending$amount, 2), ")"))
print(paste("Region with lowest spending:", region_with_lowest_spending$region, "-> ($", round(region_with_lowest_spending$amount, 2), ")"))
```




**DATA VISUALIZATION**

**1**
```{r}
# Bar plot
barplot(avg_transaction_by_region$amount, names.arg = avg_transaction_by_region$region,
        xlab = "Region", ylab = "Average Transaction Amount", main = "Average Transaction Amount by Region")

```


# 2
**Regression Analysis**
```{r}
# Step 1: Fit a Regression Model
model <- lm(amount ~ items, data = new_Demographic_Data)

# Step 2: Assess Model Fit
summary(model)  # View summary statistics of the regression model

# Step 3: Interpret Model Coefficients
coef(model)  # View coefficients of the regression model

```
**The analysis shows that there is no significant relationship between the number of items purchased and the amount spent. The p-value (0.917) and the near-zero R-squared value indicate that the number of items bought does not meaningfully impact the transaction amount. The data does not support a relationship between the number of items purchased and how much is spent.**
**The analysis shows no significant relationship between the number of items purchased and the amount spent.**


# 3
**Age Distribution by Region**
**Differences in Age by Region**
```{r}
# Summary statistics: mean age by region
summary_by_region <- aggregate(age ~ region, data = new_Demographic_Data, FUN = mean)
summary_by_region

```
**This show that there are differences in the age of customers between regions with the South region having the highest** 

**DATA VISUALIZATION**

**2**
```{r}
# Bar plot of mean age by region
barplot(summary_by_region$age, names.arg = summary_by_region$region,
        xlab = "Region", ylab = "Mean Age", main = "Mean Age by Region", col = "skyblue")

# Box plot
boxplot(age ~ region, data = new_Demographic_Data,
        xlab = "Region", ylab = "Age", main = "Age Distribution by Region")

# Violin plot
library(ggplot2)
ggplot(new_Demographic_Data, aes(x = region, y = age)) +
  geom_violin(trim = FALSE) +
  xlab("Region") +
  ylab("Age") +
  ggtitle("Age Distribution by Region")
```


```{r}
#Analysis_of_Variance_result <- aov(age ~ region, data = new_Demographic_Data)
#summary(Analysis_of_Variance_result)
```

**Predict Customer Age Based on Demographic Data**
```{r}
# Fit a linear regression model
model <- lm(age ~ region, data = new_Demographic_Data)
summary(model)

# Predict age using the model
predicted_age <- predict(model, newdata = new_Demographic_Data)
new_Demographic_Data$predicted_age <- predicted_age
```

# 4
## Correlation Between Age and Transaction Type
```{r}
# Calculate correlation between Age and Transaction_Type(in-store or online)
correlation_age_transaction_type <- cor(Demographic_Data$age, Demographic_Data$in.store)
print(paste("Correlation between Age and Transaction Type:", correlation_age_transaction_type))
```
**The correlation coefficient value of -0.178 is relatively small in magnitude. Correlation coefficients range from -1 to 1, where values closer to -1 or 1 indicate stronger relationships, and values closer to 0 indicate weaker relationships. In this case, the value is close to 0, suggesting a weak correlation.**
**The negative sign indicates that there is an inverse relationship between age and the likelihood of a transaction being made online. Specifically, as age increases, the likelihood of making an online purchase decreases slightly, and vice versa.**


```{r}
#Changing data type
Demographic_Data_A$in.store <- as.factor(Demographic_Data_A$in.store)
Demographic_Data_A$age <- as.numeric(Demographic_Data_A$age)
Demographic_Data_A$amount <- as.numeric(Demographic_Data_A$amount)
```

## Box Plot Visualizations
```{r}
# Box plot of Age by Transaction Type
ggplot(Demographic_Data_A, aes(x = in.store, y = age)) +
  geom_boxplot() +
  xlab("Transaction Type") +
  ylab("Age") +
  ggtitle("Age Distribution by Transaction Type")

# Box plot of Transaction Amount by Transaction Type
ggplot(Demographic_Data_A, aes(x = in.store, y = amount)) +
  geom_boxplot() +
  xlab("Transaction Type") +
  ylab("Transaction Amount") +
  ggtitle("Transaction Amount by Transaction Type")
```

# 5

## Model Building and Evaluation

**Preparing the Data**
```{r}
# Ensuring the 'Region' column is a factor
Demographic_Data_B$region <- as.factor(Demographic_Data_B$region)

# Handling missing values (if any)
Demographic_Data_B <- na.omit(Demographic_Data_B)

# Encoding categorical variables 
Demographic_Data_B$in.store <- as.factor(Demographic_Data_B$in.store)

```

**Splitting the Data**
```{r}
# Splitting the Data into Training and Testing Sets
set.seed(123) # for reproducibility
trainIndex <- createDataPartition(Demographic_Data_B$region, p = 0.8, list = FALSE)
trainData <- Demographic_Data_B[trainIndex,]
testData <- Demographic_Data_B[-trainIndex,]
```

**Training the Model**
```{r}
# Training a Random Forest Model
model <- randomForest(region ~ ., data = trainData)
```

**Evaluating the Model**
```{r}
#Evaluating the Model
predictions <- predict(model, newdata = testData)
confusionMatrix(predictions, testData$region)
```

**Predicting on New Data**
```{r}
new_data_predictions <- predict(model, newdata = Demographic_Data_B)
# View the predicted values

```

```{r}
# Add predictions as a new column in the Demographic_Data_B dataframe
Demographic_Data_B$Predicted_Region <- new_data_predictions

# View the updated dataframe with predictions
head(Demographic_Data_B)
```


